Inequality in socioemotional skills: a cross-cohort comparison

### 2018-10-01 codes/R/paper/text/paper_jpube_2datamethods.tex
fix bibliography

A common approach to infer a child's socio-emotional development is based on behavioural screening scales. As part of these tools, mothers (or teachers) indicate whether their children exhibit a series of behaviours -- the \emph{items} of the scale. In the British and Millennium Cohort Studies, two different scales were employed. In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS cohort, mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}). The SDQ was created as an update to the Rutter scale. It encompasses more recent advances in child psychopathology, and emphasises positive traits alongside undesirable ones \citep{Stone2010}. The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively. In the child psychiatry and psychology literatures, the Rutter and SDQ behavioural screening scales are regarded as measures of behavioural problems and mental health. However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{goodman2011population}.

@@

The original Rutter scale, used for the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}. This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \cite{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}. However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used for the MCS cohort, was conceived to have five subscales of five items each. The five subscales are: \emph{hyperactivity}, \emph{emotional symptoms}, \emph{conduct problems}, \emph{peer problems}, and \emph{prosocial}. This five-factor structure has been validated in many contexts \citep{Stone2010}; lower-dimensional structures have been also suggested \citep{Dickey2004}. Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}.

We use exploratory factor analysis (EFA) to assess the factors structure of our 11-item scale combining Rutter and SDQ.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and now also used in the economics literature. The results are displayed in \autoref{tab:numfac}. As pointed out in \citet{conti2014bayesian}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors in the MCS. In our analysis, we adopt two factors and a dedicated measurement system, where each measure reflects only one factor. This choice is justified both by the child psychology literature cited above, and as compromise to work with the same number of factors in the two cohorts. The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very smilar. Following previous research, we name the first dimension \emph{Externalising skills} (EXT, comprising the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, comprising the items worried, fearful, solitary, unhappy, and aches).

@@

We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,conti2010education,Attanasio2017c}. We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}

@@

The model restrictions in \eqref{eq:propens} and \eqref{eq:thresh} do not identify the parameters without additional assumptions. As per the traditional factor analysis approach, we impose a normal distribution on the latent skills and error terms:\footnote{Recent work has also used mixtures of normals for the latent factors distribution, e.g. \citet{conti2010education}.}
\be\label{eq:norm}
\bm{\theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c).
\ee

@@

In the framework of factor analysis, measurement invariance is a formally testable property. In this paper, we follow the recent identification methodology by \cite{Wu2016a}. The configural model defined in the previous section serves as the starting point. Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models. These models place increasing restrictions on the item parameters, constraining them to be equal across groups. Their fit is then compared to that of the configural model. Intuitively, if the additional cross-group restrictions have not significantly worsened model fit, one can conclude that a certain level of invariance is achieved. The hierarchy of restrictions is detailed in \autoref{tab:invparam}.

@@

Comparison of $\chi^2$ values across models is a common likelihood-based strategy. However, tests based on $\Delta \chi^2$ are known to display high Type I error rates with large sample size and more complex models such as our own \citep{Sass2014}. In fact, for all invariance levels in our applications a chi-squared difference would point to a lack of measurement invariance. The use of approximate fit indices (AFIs) is therefore recommended alongside $\chi^2$. These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance. Nevertheless, AFIs are widely used in empirical practice to assess model fit.\footnote{The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice. Simulation evidence by \cite{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$. \cite{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \cite{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}

@@

This is an important finding, which has to our knowledge never been acknowledged in this literature. How can this lack of comparability be explained?  A possible interpretation is connected with secular evolution of social and cultural norms about child behaviours. For example, commonly held views of what constitutes a restless, distracted, or unhappy child might have changed between 1975 and 2005/6.\footnote{Calibrating the Rutter and SDQ using a contemporary sample of children cannot rule out this issue. For example, \cite{Collishaw2004} administered both Rutter and SDQ items to parents of a small sample of adolescents in London. They use the mapping between the two questionnaires to impute Rutter scores for mothers who answered the SDQ. This can correct for contemporaneous reporting differences between questionnaires, but cannot tackle reporting differences between samples collected at different times in history.}

>>>>

A common approach to infer a child's socio-emotional development is based on behavioural screening scales. As part of these tools, mothers (or teachers) indicate whether their children exhibit a series of behaviours -- the \emph{items} of the scale. In the British and Millennium Cohort Studies, two different scales were employed. In the BCS, the Rutter A Scale was used \citep{Rutter1970} while in the MCS cohort, mothers were  administered the Strengths and Difficulties Questionnaire (SDQ, \citealp{Goodman1994,Goodman1997}). The SDQ was created as an update to the Rutter scale. It encompasses more recent advances in child psychopathology, and emphasises positive traits alongside undesirable ones \citep{Stone2010}. The Rutter and SDQ scales are reproduced in \autoref{tab:scales}; they have 23 and 25 items each, respectively. In the child psychiatry and psychology literatures, the Rutter and SDQ behavioural screening scales are regarded as measures of behavioural problems and mental health. However, in our analysis we follow the economics literature, and - after having recoded them accordingly - we interpret them as measures of positive child development \citep{Goodman2011}.

@@

The original Rutter scale, used for the BCS cohort, distinguishes behaviours into two subscales: \emph{anti-social} and \emph{neurotic} \citep{Rutter1970}. This two-factor conceptualisation has been validated using data from multiple contexts, and the latent dimensions have been broadly identified as externalising and internalising behaviour problems.\footnote{See for example \citet{Fowler1979,Venables1983,Tremblay1987,Berglund1999,Klein2009}. However, in some cases a three-factor structure was found to better fit the data, with the externalising factor separating into two factors seemingly capturing aggressive and hyperactive behaviours \citep{Behar1974,McGee1985}.} The Strength and Difficulties Questionnaire, used for the MCS cohort, was conceived to have five subscales of five items each. The five subscales are: \emph{hyperactivity}, \emph{emotional symptoms}, \emph{conduct problems}, \emph{peer problems}, and \emph{prosocial}. This five-factor structure has been validated in many contexts \citep{Stone2010}; lower-dimensional structures have been also suggested \citep{Dickey2004}. Recent research has shown that there are some benefits to using broader subscales that correspond to the externalising and internalising factors in Rutter, especially in low-risk or general population samples \citep{Goodman2010}.

We use exploratory factor analysis (EFA) to assess the factors structure of our 11-item scale combining Rutter and SDQ.\footnote{Factor-analytic methods have long been used in psychology, and in recent years they have become increasingly popular in economics, especially to meaningfully aggregate high-dimensional items measuring different aspects of common underlying dimensions of human development. The EFA is performed decomposing the polychoric correlation matrix of the items and using weighted least squares, and the solution is rescaled using oblique factor rotation (\emph{oblimin}). We use the \textbf{\textsf{R}} package \texttt{psych}, version 1.8.4 \citep{Revelle2018}.} We start by investigating the number of latent constructs that are captured by the scale, using different methods developed in the psychometric literature, and now also used in the economics literature. The results are displayed in \autoref{tab:numfac}. As pointed out in \citet{Conti2014}, there is relatively little agreement among procedures; this is the case especially for the Rutter items in the BCS data, where different methods suggest to retain between 1 and 3 factors, while most methods suggest to retain 2 factors in the MCS. In our analysis, we adopt two factors and a dedicated measurement system, where each measure reflects only one factor. This choice is justified both by the child psychology literature cited above, and as compromise to work with the same number of factors in the two cohorts. The two-factor EFA delivers a neat and sensible separation between items, as shown in \autoref{tab:load}: reassuringly, similarly-worded items load on the same factor across the two cohorts, and also the magnitude of the respective loadings (measuring the strength of the association between the item and the factor) is very smilar. Following previous research, we name the first dimension \emph{Externalising skills} (EXT, comprising the items restless, squirmy/fidgety, fights/bullies, distracted, tantrums, and disobedient) and the second dimension \emph{Internalising skills} (INT, comprising the items worried, fearful, solitary, unhappy, and aches).

@@

We make the common assumption of a dedicated (or congeneric) factor structure, where each measure is assumed to load on only one latent dimension \citep{Heckman2013,Conti2010a,Attanasio2018a}. We mirror the structure found in the exploratory factor analysis (see \autoref{tab:load}), and assume that items 1-6 load exclusively on the externalising factor and items 7-11 on the internalising factor.\footnote{The dedicated factor structure corresponds to a sparse loading matrix, i.e.: $$\bm{\Lambda}_{c} \coloneqq \begin{bmatrix} \lambda_{1c}, \dots, \lambda_{6c} & \bm{0} \\ \bm{0} & \lambda_{7c}, \dots, \lambda_{11c} \end{bmatrix}.$$}

@@

The model restrictions in \eqref{eq:propens} and \eqref{eq:thresh} do not identify the parameters without additional assumptions. As per the traditional factor analysis approach, we impose a normal distribution on the latent skills and error terms:\footnote{Recent work has also used mixtures of normals for the latent factors distribution, e.g. \citet{Conti2010a}.}
\be\label{eq:norm}
\bm{\theta}_{jc} \sim N(\bm{\kappa}_c, \bm{\Phi}_c) \qquad \text{and} \qquad \bm{u}_{jc} \sim N(\bm{0}, \bm{\Psi}_c).
\ee

@@

In the framework of factor analysis, measurement invariance is a formally testable property. In this paper, we follow the recent identification methodology by \citet{Wu2016a}. The configural model defined in the previous section serves as the starting point. Measurement invariance is then assessed by comparing the configural model to a series of hierarchically nested models. These models place increasing restrictions on the item parameters, constraining them to be equal across groups. Their fit is then compared to that of the configural model. Intuitively, if the additional cross-group restrictions have not significantly worsened model fit, one can conclude that a certain level of invariance is achieved. The hierarchy of restrictions is detailed in \autoref{tab:invparam}.

@@

Comparison of $\chi^2$ values across models is a common likelihood-based strategy. However, tests based on $\Delta \chi^2$ are known to display high Type I error rates with large sample size and more complex models such as our own \citep{Sass2014}. In fact, for all invariance levels in our applications a chi-squared difference would point to a lack of measurement invariance. The use of approximate fit indices (AFIs) is therefore recommended alongside $\chi^2$. These indices do not have a known sampling distribution, thus making it necessary to rely on rules of thumb to assess what level of \textDelta AFI indicates invariance. Nevertheless, AFIs are widely used in empirical practice to assess model fit.\footnote{The root mean squared error of approximation (RMSE) and the Tucker-Lewis index (TLI) are traditionally the most used AFIs in empirical practice. Simulation evidence by \citet{Cheung2002} shows that these indices can show correlation between overall and relative fit, and suggest relying on additional indices, such as the comparative fit index (CFI, \citealp{Bentler1990}), McDonald non-centrality index (MFI, \citealp{McDonald1989}), and Gamma-hat index \citep{Steiger1989} for the case of ordered measures. Commonly accepted thresholds for rejection are $\Delta CFI < - 0.01$, $\Delta MFI < - 0.02$, and $\Delta Gamma hat < - 0.001$. \citet{Meade2008}, using the results from a simulation study, suggests stricter thresholds that should apply in a variety of conditions. For CFI, a single cutoff value of $.002$ is proposed, while cutoffs for MFI depend on the problem's characteristics; in our case (2 factors, 11 items), they suggest $.0066$. \citet{Sass2014} however cast some doubts of the generalisability of these cutoffs to WLSMV estimators.}

@@

This is an important finding, which has to our knowledge never been acknowledged in this literature. How can this lack of comparability be explained?  A possible interpretation is connected with secular evolution of social and cultural norms about child behaviours. For example, commonly held views of what constitutes a restless, distracted, or unhappy child might have changed between 1975 and 2005/6.\footnote{Calibrating the Rutter and SDQ using a contemporary sample of children cannot rule out this issue. For example, \citet{Collishaw2004} administered both Rutter and SDQ items to parents of a small sample of adolescents in London. They use the mapping between the two questionnaires to impute Rutter scores for mothers who answered the SDQ. This can correct for contemporaneous reporting differences between questionnaires, but cannot tackle reporting differences between samples collected at different times in history.}

### 2018-10-01 codes/R/paper/text/paper_jpube_3resultsconclusion.tex
fix bibliography

Parameter estimates from our factor model are presented in \autoref{tab:measpars}. As discussed in the previous section, loadings and thresholds are constrained to have the same value across groups. Intercepts are normalised to zero, and error variances to one, for the reference group -- males in the BCS cohort. We use the estimates from this model to predict a score for each child in our sample along the latent externalising and internalising socio-emotional skill dimensions.\footnote{We use an empirical Bayes modal (EBM) approach to estimate the scores. The parameters are estimated using three sources of information. The first is the distribution of the latent variables $\bm{\theta}$, treated as random parameters with a prior $h(\bm{\theta}, \bm{\Omega})$, conditional on the parameters $\bm{\Omega}$. This prior is assumed to be multivariate normal. The second is the observed data $\bm{X}$, and the third is the estimated parameters $\hat{\bm{\Omega}}$. Data and prior are combined into the posterior distribution $w(\bm{\theta}| \bm{X}, \hat{\bm{\Omega}})$. For further details, see Chapter 7 in \cite{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}. The unit of measurement is standard deviations of the distribution in the subsample of males in the BCS. Given our measurement invariance results in section \ref{sec:measinv}, we stress that the \emph{location} of these scores should not be directly compared across cohorts. However, the shape of the distribution can be given a cross-cohort interpretation.\footnote{The density of the scored factors can be contrasted with the distribution of sum scores in \autoref{fig:sumscores}. Using raw scores, instead, shows an increase in mass only at the top of the distribution.} It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.

@@

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{gaysina2013maternal}; less evidence there is however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{reardon2016recent} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.

@@

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males. This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{autor2016family}. Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS. Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{roantree2018women}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{blundell2007changes}.

@@

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, he is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{goodman2015social}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{hsieh2018smoking} have shown personality to be a key mechanism through which peers affect smoking behaviour.

Conditional on socio-emotional skills, cognition has limited predictive power for these behaviours, and only for girls.\footnote{We do not observe significant associations between early socio-emotional skills and other risky behaviours like drug-taking and alcohol consumption. One possible reason might be the relatively young age at which these skills are measured. Results are available upon request.} This is in line with the evidence in \citet{conti2010understanding}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study. \citet{conti2013personality} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI). We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.

For the British Cohort Study, we last examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition. This is consistent with the evidence in \citet{conti2011early}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.

@@

In this paper we have studied inequality in a dimension of human capital which has received limited attention in the literature so far: socio-emotional skills very early in life. In particular, we have focused on the measurements of these skills at age 5 in two British cohorts born 30 years apart: the one of children born in 1970 (British Cohort Study, BCS) and the one of children born in 2000/1 (Millenium Cohort Study, MCS). We have provided several contributions to the recent but flourishing literature on the determinants and consequences of early human development.

>>>>

Parameter estimates from our factor model are presented in \autoref{tab:measpars}. As discussed in the previous section, loadings and thresholds are constrained to have the same value across groups. Intercepts are normalised to zero, and error variances to one, for the reference group -- males in the BCS cohort. We use the estimates from this model to predict a score for each child in our sample along the latent externalising and internalising socio-emotional skill dimensions.\footnote{We use an empirical Bayes modal (EBM) approach to estimate the scores. The parameters are estimated using three sources of information. The first is the distribution of the latent variables $\bm{\theta}$, treated as random parameters with a prior $h(\bm{\theta}, \bm{\Omega})$, conditional on the parameters $\bm{\Omega}$. This prior is assumed to be multivariate normal. The second is the observed data $\bm{X}$, and the third is the estimated parameters $\hat{\bm{\Omega}}$. Data and prior are combined into the posterior distribution $w(\bm{\theta}| \bm{X}, \hat{\bm{\Omega}})$. For further details, see Chapter 7 in \citet{Skrondal2004}.} We plot the distribution of the scores in \autoref{fig:dens}. The unit of measurement is standard deviations of the distribution in the subsample of males in the BCS. Given our measurement invariance results in section \ref{sec:measinv}, we stress that the \emph{location} of these scores should not be directly compared across cohorts. However, the shape of the distribution can be given a cross-cohort interpretation.\footnote{The density of the scored factors can be contrasted with the distribution of sum scores in \autoref{fig:sumscores}. Using raw scores, instead, shows an increase in mass only at the top of the distribution.} It is immediately visible that there is more mass in the tails of the distribution in the 2000 than in the 1970 cohort.

@@

\autoref{fig:ineqsm} shows an even starker pattern when comparing children of mothers who smoked in pregnancy with non-smoking mothers. The fact that maternal smoking during pregnancy is a risk factor for offspring behavioural problems is well known in the medical literature \citep{Gaysina2013}; less evidence there is however, on whether and to which extent these associations have changed across cohorts. The difference in child skills has increased, from less than .2 to around .4 of a standard deviation, again with the biggest increase experienced by the boys. There is also a significant increase in the gradient by paternal occupation based on social class (\autoref{fig:ineqfsc}), although this is less pronounced if compared to the one based on maternal characteristics. In particular, male children with no father figure living in their household have worse skills compared to children with blue collar fathers in the MCS cohort. Otherwise, skill differences in father's occupation are mostly constant across the two cohorts.\footnote{Figures \ref{fig:ineqips}, \ref{fig:ineqism}, and \ref{fig:ineqifsc} show inequality in the scale items underlying the factor scores used in this section. The increase in inequality across cohorts is still present, but less marked when looking at these single items. This shows the importance of the factor analysis step in aggregating items, explicitly modelling the measurement error, and testing and accounting for (loadings and thresholds) invariance across the two cohorts.} These patterns are in stark contrast with the findings of \citet{Reardon2016} for the US, who have found a narrowing of the readiness gaps from 1998 to 2010.

@@

Overall, the importance of maternal socioeconomic status (education and in particular employment) in determining socio-emotional skills has increased from the BCS to the MCS children. The `premium' in skills for children of better educated and employed mothers is significantly larger, for both boys and girls, internalising and externalising skills. At the same time, the penalty for having a blue-collar father, or not having a father figure at all in the household, has significantly declined across the two cohorts, especially for girls. Being born to an unmarried mother, and to a mother who smoked during pregnancy, is associated with a higher penalty for both dimensions of socio-emotional skills in the latter cohort, but only for males. This is consistent with recent evidence which shows that family disadvantage disproportionately impedes the pre-market development of boys, in terms of higher disciplinary problems, lower achievement scores, and fewer high-school completions \citep{Autor2016a}. Girls of non-white ethnicity, instead, have worse internalising and externalising skills in the MCS, a penalty not suffered by 5-year old non-white girls in the BCS. Firstborn boys and girls in the BCS have worse skills, but this difference disappears in the MCS. Lastly, we document an increase in the returns to birth weight, which is more pronounced for boys' internalising skills.

These changes in the relative importance of pregnancy factors and family background characteristics for child socio-emotional skills at age 5 need to be interpreted in the light of the significant changes in the prevalence of such characteristics across cohorts. As shown in \autoref{tab:sumstats}, the age of the mother at birth, proportion of mothers non-smoking in pregnancy, with post-compulsory education and in employment at the age 5 of the child has substantially increased; at the same time, the proportion of households with no father figure has increased, and so the proportion of women unmarried at birth is much higher in the 2000 than in the 1970 cohort. Also, the ethnic structure of the population has changed, with a higher proportion of non-white children in the MCS than in the BCS. In general, this has been a period of significant societal changes, with an almost continual rise in the proportion of women in employment, an older age at first birth and a rise in dual-earning parents families \citep{Roantree2018}. However, here we do not attempt to disentangle whether and to which extent the observed changes in inequality in socio-emotional skills across the two cohorts can be attributed to changes in returns (or penalties) to maternal characteristics (such as education and employment) or to compositional changes, like it has been done for the analysis of wage inequality \citep{Blundell2007}.

@@

Socio-emotional skills at five years of age are predictive of adolescent health behaviour and outcomes in both cohorts.\footnote{Unfortunately the strength of the association cannot be directly compared, since the outcomes are measured at different ages: 16 and 14 years for BCS and MCS, respectively.} \autoref{tab:adol} examines adolescent smoking and BMI for both cohorts; \autoref{tab:bcs_adultapp} reports the results for the same outcomes in adulthood (at age 42), for the BCS only. Externalising skills are negatively correlated to subsequent smoking and BMI in both cohorts, for both genders. Recall that a child with high externalising skills exhibits less restless and hyperactive behaviours, and has less anti-social conduct. Our findings are consistent with the body of evidence reviewed in section \ref{sec:lit}, which shows that better socio-emotional skills (measured using different scales and at various points during childhood and adolescence) are negatively associated with smoking. At the same time, internalising skills are positively correlated with smoking (only in the 1970 cohort) and BMI (only for girls), although less strongly than externalising skills. This apparently counterintuitive result makes sense in light of the items in our internalising scale shown in \autoref{tab:scalecomp}. A child with better internalising skills is less solitary, neurotic, and worried. From this perspective, he is likely more sociable and subject to peer influence in health behaviours. This is consistent with the evidence in \citet{Goodman2015}, who find a positive association between child emotional health (measured with items for the internalizing behaviour subscale of the Rutter scale at age 10 in the BCS) and smoking at age 42. Furthermore, in recent work \citet{Hsieh2018} have shown personality to be a key mechanism through which peers affect smoking behaviour.

Conditional on socio-emotional skills, cognition has limited predictive power for these behaviours, and only for girls.\footnote{We do not observe significant associations between early socio-emotional skills and other risky behaviours like drug-taking and alcohol consumption. One possible reason might be the relatively young age at which these skills are measured. Results are available upon request.} This is in line with the evidence in \citet{Conti2010}, who show that not accounting for non-cognitive traits (a self-regulation factor measured at age 10) overestimates the importance of cognition for predicting health and health behaviours, using data from the British cohort study. \citet{Conti2013a} use rich data on child personality and socio-emotional traits collected at ages 7, 11 and 16 in the 1958 British birth cohort,\footnote{They use the Rutter scale and the Bristol Social Adjustment Guide.} and show that these traits rival the importance of cognition in explaining the education gradient in health behaviours (including smoking and BMI). We show that child socio-emotional skills have greater predictive power for health outcomes and behaviours even when measured at an earlier age.

For the British Cohort Study, we last examine the association between socio-emotional skills at age five and adult education and labour market outcomes. The structure of \autoref{tab:bcs_adult} is similar to \autoref{tab:adol}, but it considers educational achievement, employment, and earnings (conditional on being in paid employment) for the BCS cohort members. For these outcomes, the predictive power of cognitive skills outweighs that of socio-emotional skills, whose predictive power diminishes over time (between the ages 34 and 42), and is driven to insignificance after controlling for cognition. This is consistent with the evidence in \citet{Conti2011}, who show that cognitive endowments at age 10 are more predictive (than socio-emotional and health ones) for employment and wage outcomes in the BCS. Again, we show that the greater predictive power of cognition for socioeconomic outcomes holds even when considering earlier-life measures of child development.

@@

In this paper we have studied inequality in a dimension of human capital which has received limited attention in the literature so far: socio-emotional skills very early in life. In particular, we have focused on the measurements of these skills at age 5 in two British cohorts born 30 years apart: the one of children born in 1970 (British Cohort Study, BCS) and the one of children born in 2000/1 (Millennium Cohort Study, MCS). We have provided several contributions to the recent but flourishing literature on the determinants and consequences of early human development.
